---
title: "General Visium HD Data Analysis Workflow"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 8)

# Set library paths - adjust according to your environment
.libPaths(c("/sc/arion/projects/naiklab/ikjot/R/",
           "/hpc/packages/minerva-rocky9/rpackages/4.4.1/site-library",
           "~/R/x86_64-pc-linux-gnu-library/4.4/")) 

# Load required libraries
library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)
library(RColorBrewer)
library(ggpubr)

# Set up parallel processing
future::plan("multicore", workers = future::availableCores())
options('future.globals.maxSize' = 2*10^12)

# Create output directories
dir.create("output", showWarnings = FALSE)
dir.create("output/plots", showWarnings = FALSE, recursive = TRUE)
dir.create("output/tables", showWarnings = FALSE, recursive = TRUE)
dir.create("output/RDS-files", showWarnings = FALSE, recursive = TRUE)
```

# Visium HD Data Analysis Workflow

This is a comprehensive workflow for analyzing Visium HD spatial transcriptomics data. The workflow includes data loading, quality control, normalization, clustering, and spatial analysis.

## 1. Data Loading and Initial Setup

### 1.1 Configure Analysis Parameters

```{r parameters}
# =============================================================================
# ANALYSIS PARAMETERS - MODIFY THESE FOR YOUR DATASET
# =============================================================================

# Sample information
sample.names <- c("SAMPLE1", "SAMPLE2")  # Update with your sample names
data.base.dir <- "data/raw/"  # Update with your data directory path
bin.size <- c(2)  # Bin size for analysis (2um, 8um, etc.)

# Image names for each sample (if multiple images per sample)
image.names <- list(
  "SAMPLE1" = c("IMAGE1", "IMAGE2"),  # Update with your image names
  "SAMPLE2" = c("IMAGE3", "IMAGE4")
)

# QC thresholds
min.counts <- 50      # Minimum UMI counts per spot
max.mt.percent <- 20  # Maximum mitochondrial gene percentage

# Clustering parameters
sketch.ncells <- 50000  # Number of cells for sketching
cluster.resolutions <- c(0.2, 0.5, 1, 2)  # Clustering resolutions to test

# Spatial analysis parameters
banksy.lambda <- c(0.2, 0.8)  # BANKSY lambda values
banksy.k <- c(15, 20)          # BANKSY k-neighbors
```

### 1.2 Load Visium HD Data

```{r load_data}
# Function to load Visium HD data
load_visium_hd_data <- function(sample.name, data.dir, bin.size, image.names) {
  sample.data.dir <- file.path(data.dir, sample.name)
  
  if (file.exists(sample.data.dir)) {
    object <- Load10X_Spatial(
      data.dir = sample.data.dir,
      bin.size = bin.size,
      image = image.names[[sample.name]]
    )
    cat("Loaded sample:", sample.name, "with", ncol(object), "spots\n")
    return(object)
  } else {
    warning(paste("Directory not found:", sample.data.dir))
    return(NULL)
  }
}

# Load all samples
object.list <- lapply(sample.names, function(sample.name) {
  load_visium_hd_data(sample.name, data.base.dir, bin.size, image.names)
})
names(object.list) <- sample.names

# Remove NULL objects (failed loads)
object.list <- object.list[!sapply(object.list, is.null)]

cat("Successfully loaded", length(object.list), "samples\n")
```

### 1.3 Split Multi-Image Samples (if applicable)

```{r split_samples}
# Function to split samples with multiple images
split_multi_image_samples <- function(object.list, image.names) {
  split.object.list <- list()
  
  for (sample.name in names(object.list)) {
    obj <- object.list[[sample.name]]
    images <- image.names[[sample.name]]
    
    if (length(images) > 1) {
      # If CSV files exist for image splitting
      csv.dir <- file.path(dirname(obj@images[[1]]@image), "cloupe_csvs")
      
      if (dir.exists(csv.dir)) {
        csv.files <- list.files(csv.dir, pattern = "\\.csv$", full.names = TRUE)
        
        for (i in seq_along(images)) {
          image.name <- images[i]
          csv.file <- csv.files[grep(image.name, csv.files)][1]
          
          if (!is.na(csv.file) && file.exists(csv.file)) {
            cells.to.keep <- read.csv(csv.file)$Barcode
            sub.obj <- subset(obj, cells = cells.to.keep)
            sub.obj$orig.ident <- image.name
            sub.obj$sample <- sample.name
            
            split.object.list[[paste(sample.name, image.name, sep = "_")]] <- sub.obj
          }
        }
      } else {
        # If no CSV files, keep the whole sample
        obj$orig.ident <- sample.name
        obj$sample <- sample.name
        split.object.list[[sample.name]] <- obj
      }
    } else {
      # Single image sample
      obj$orig.ident <- images[1]
      obj$sample <- sample.name
      split.object.list[[sample.name]] <- obj
    }
  }
  
  return(split.object.list)
}

# Split samples if needed
split.object.list <- split_multi_image_samples(object.list, image.names)
cat("After splitting:", length(split.object.list), "objects\n")
```

## 2. Quality Control Analysis

### 2.1 Calculate QC Metrics

```{r qc_metrics}
# Calculate mitochondrial gene percentage
split.object.list <- lapply(split.object.list, function(obj) {
  obj <- PercentageFeatureSet(obj, pattern = "^mt-|^MT-", col.name = "percent.mt")
  return(obj)
})

# Get assay name dynamically
assay.name <- names(split.object.list[[1]]@assays)[1]
count.col <- paste0("nCount_", assay.name)
feature.col <- paste0("nFeature_", assay.name)

cat("Using assay:", assay.name, "\n")
cat("Count column:", count.col, "\n")
cat("Feature column:", feature.col, "\n")
```

### 2.2 QC Visualization

```{r qc_plots, fig.height=12, fig.width=14}
# Combine metadata for visualization
merged.md <- bind_rows(
  lapply(split.object.list, function(obj) {
    return(obj@meta.data)
  }), 
  .id = "sampleName"
)

# QC violin plots
p1 <- merged.md %>%
  ggplot(aes_string(x = "sampleName", y = paste0("log10(", count.col, ")"), fill = "sampleName")) +
  geom_violin() +
  theme_classic() +
  geom_hline(yintercept = log10(min.counts), lty = 2, color = "red") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("UMI Counts per Spot")

p2 <- merged.md %>%
  ggplot(aes_string(x = "sampleName", y = paste0("log10(", feature.col, ")"), fill = "sampleName")) +
  geom_violin() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Features per Spot")

p3 <- merged.md %>%
  ggplot(aes(x = sampleName, y = percent.mt, fill = sampleName)) +
  geom_violin() +
  theme_classic() +
  geom_hline(yintercept = max.mt.percent, lty = 2, color = "red") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Mitochondrial Gene %")

p4 <- merged.md %>%
  ggplot(aes_string(x = "sampleName", 
                   y = paste0(count.col, "/", feature.col), 
                   fill = "sampleName")) +
  geom_violin() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("UMI per Feature Ratio")

# Density plots
p5 <- merged.md %>%
  ggplot(aes_string(x = paste0("log10(", count.col, ")"), fill = "sampleName")) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = log10(min.counts), color = "red", linetype = "dashed") +
  theme_classic() +
  ggtitle("UMI Count Distribution")

p6 <- merged.md %>%
  ggplot(aes_string(x = paste0("log10(", feature.col, ")"), fill = "sampleName")) +
  geom_density(alpha = 0.5) +
  theme_classic() +
  ggtitle("Feature Count Distribution")

# Combine plots
(p1 | p2) / (p3 | p4) / (p5 | p6)

# Spatial QC plots
pdf("output/plots/spatial_qc_plots.pdf", height = 8, width = 12)
for (obj.name in names(split.object.list)) {
  obj <- split.object.list[[obj.name]]
  
  p1 <- SpatialFeaturePlot(obj, features = count.col) + 
    ggtitle(paste(obj.name, "- UMI Counts"))
  p2 <- SpatialFeaturePlot(obj, features = feature.col) + 
    ggtitle(paste(obj.name, "- Feature Counts"))
  p3 <- SpatialFeaturePlot(obj, features = "percent.mt") + 
    ggtitle(paste(obj.name, "- Mitochondrial %"))
  
  print((p1 | p2) / p3)
}
dev.off()
```

### 2.3 Apply QC Filters

```{r qc_filter}
# Apply QC filters
filtered.object.list <- lapply(split.object.list, function(obj) {
  cat("Before filtering:", ncol(obj), "spots\n")
  
  # Apply filters
  filter_expr <- substitute(
    nCount >= min.counts & percent.mt <= max.mt.percent,
    list(
      nCount = as.name(count.col),
      min.counts = min.counts,
      max.mt.percent = max.mt.percent
    )
  )
  
  obj <- subset(obj, subset = eval(filter_expr))
  cat("After filtering:", ncol(obj), "spots\n")
  
  return(obj)
})

# Summary of filtering
cat("\nFiltering Summary:\n")
for (obj.name in names(filtered.object.list)) {
  original <- ncol(split.object.list[[obj.name]])
  filtered <- ncol(filtered.object.list[[obj.name]])
  cat(obj.name, ":", original, "->", filtered, 
      "(", round(100 * filtered/original, 1), "%)\n")
}
```

## 3. Data Integration and Normalization

### 3.1 Normalize and Merge Data

```{r normalize_merge}
# Normalize each sample individually
normalized.object.list <- lapply(filtered.object.list, function(obj) {
  obj <- NormalizeData(obj, assay = assay.name)
  return(obj)
})

# Merge all samples
if (length(normalized.object.list) > 1) {
  merged.object <- merge(normalized.object.list[[1]], 
                        normalized.object.list[2:length(normalized.object.list)])
} else {
  merged.object <- normalized.object.list[[1]]
}

cat("Merged object dimensions:", dim(merged.object), "\n")
```

### 3.2 Sketching for Large Datasets

```{r sketching}
# Set default assay
DefaultAssay(merged.object) <- assay.name

# Find variable features
merged.object <- FindVariableFeatures(merged.object)
merged.object <- ScaleData(merged.object)

# Create sketch for large datasets
merged.object <- SketchData(
  object = merged.object,
  ncells = sketch.ncells,
  method = "LeverageScore",
  sketched.assay = "sketch"
)

cat("Sketched", sketch.ncells, "cells for downstream analysis\n")
```

## 4. Dimensionality Reduction and Clustering

### 4.1 PCA and Clustering on Sketched Data

```{r clustering}
# Switch to sketched assay
DefaultAssay(merged.object) <- "sketch"

# Perform clustering workflow on sketched data
merged.object <- FindVariableFeatures(merged.object)
merged.object <- ScaleData(merged.object)
merged.object <- RunPCA(merged.object, assay = "sketch", reduction.name = "pca.sketch")
merged.object <- FindNeighbors(merged.object, assay = "sketch", 
                              reduction = "pca.sketch", dims = 1:50)
merged.object <- FindClusters(merged.object, resolution = cluster.resolutions)
merged.object <- RunUMAP(merged.object, reduction = "pca.sketch", 
                        reduction.name = "umap.sketch", return.model = TRUE, dims = 1:50)
```

### 4.2 Project Results to Full Dataset

```{r projection}
# Project clustering results to full dataset
merged.object <- ProjectData(
  object = merged.object,
  assay = assay.name,
  full.reduction = "full.pca.sketch",
  sketched.assay = "sketch",
  sketched.reduction = "pca.sketch",
  umap.model = "umap.sketch",
  dims = 1:50,
  refdata = setNames(
    lapply(cluster.resolutions, function(res) paste0("sketch_snn_res.", res)),
    paste0("sketch_snn_res.", cluster.resolutions)
  )
)

# Save intermediate results
saveRDS(merged.object, file = "output/RDS-files/merged_object_clustered.RDS")
cat("Saved clustered object\n")
```

## 5. Cluster Visualization and Analysis

### 5.1 UMAP Visualization

```{r umap_viz, fig.width=16, fig.height=12}
# UMAP plots for different resolutions
DefaultAssay(merged.object) <- "sketch"
sketched.plots <- lapply(cluster.resolutions, function(res) {
  Idents(merged.object) <- paste0("sketch_snn_res.", res)
  DimPlot(merged.object, reduction = "umap.sketch", label = TRUE, 
          label.size = 3) + 
    ggtitle(paste("Sketched - Resolution", res)) + 
    theme(legend.position = "none")
})

DefaultAssay(merged.object) <- assay.name
projected.plots <- lapply(cluster.resolutions, function(res) {
  Idents(merged.object) <- paste0("sketch_snn_res.", res)
  DimPlot(merged.object, reduction = "full.umap.sketch", label = TRUE, 
          label.size = 3) + 
    ggtitle(paste("Projected - Resolution", res)) + 
    theme(legend.position = "none")
})

# Combine plots
wrap_plots(c(sketched.plots, projected.plots), ncol = 4)

# Save UMAP plots
pdf("output/plots/umap_multiple_resolutions.pdf", height = 10, width = 16)
wrap_plots(projected.plots, ncol = 2)
dev.off()
```

### 5.2 Spatial Cluster Visualization

```{r spatial_viz, fig.width=20, fig.height=16}
# Spatial plots for each resolution
pdf("output/plots/spatial_clusters_multiple_resolutions.pdf", height = 12, width = 20)
for (res in cluster.resolutions) {
  p <- SpatialDimPlot(merged.object, 
                     group.by = paste0("sketch_snn_res.", res),
                     label = TRUE, repel = TRUE, label.size = 4,
                     pt.size.factor = 2.5) +
    ggtitle(paste("Spatial Clusters - Resolution", res)) +
    theme(legend.position = "none")
  print(p)
}
dev.off()

# Individual cluster highlighting
for (res in cluster.resolutions[1:2]) {  # Limit to first two resolutions
  pdf(paste0("output/plots/individual_clusters_res_", res, ".pdf"), 
      height = 8, width = 12)
  
  Idents(merged.object) <- paste0("sketch_snn_res.", res)
  
  for (cluster in unique(Idents(merged.object))) {
    cells <- CellsByIdentities(merged.object, idents = cluster)
    
    p <- SpatialDimPlot(merged.object,
                       cells.highlight = cells,
                       cols.highlight = c("red", "grey90"),
                       pt.size.factor = 2.5) +
      ggtitle(paste("Cluster", cluster, "- Resolution", res)) +
      theme(legend.position = "none")
    print(p)
  }
  dev.off()
}
```

## 6. Marker Gene Analysis

### 6.1 Find Cluster Markers

```{r markers}
# Join layers for marker analysis
merged.object <- JoinLayers(merged.object)

# Find markers for each resolution
marker.list <- list()
for (res in cluster.resolutions) {
  cat("Finding markers for resolution", res, "\n")
  
  Idents(merged.object) <- paste0("sketch_snn_res.", res)
  markers <- FindAllMarkers(merged.object, assay = assay.name, only.pos = TRUE)
  
  marker.list[[paste0("res_", res)]] <- markers
  
  # Save markers
  write.csv(markers, 
           file = paste0("output/tables/markers_resolution_", res, ".csv"),
           row.names = FALSE)
}

# Display top markers for optimal resolution (e.g., 0.5)
optimal.res <- "0.5"
if (paste0("res_", optimal.res) %in% names(marker.list)) {
  top.markers <- marker.list[[paste0("res_", optimal.res)]] %>%
    group_by(cluster) %>%
    top_n(n = 5, wt = avg_log2FC)
  
  print(top.markers)
}
```

### 6.2 Visualize Key Markers

```{r marker_viz, fig.width=16, fig.height=12}
# Get top 2 markers per cluster for visualization
if (paste0("res_", optimal.res) %in% names(marker.list)) {
  top2.markers <- marker.list[[paste0("res_", optimal.res)]] %>%
    group_by(cluster) %>%
    top_n(n = 2, wt = avg_log2FC) %>%
    pull(gene) %>%
    unique()
  
  # Limit to first 16 markers for visualization
  markers.to.plot <- head(top2.markers, 16)
  
  # Feature plots
  pdf("output/plots/top_marker_features.pdf", height = 16, width = 20)
  print(FeaturePlot(merged.object, features = markers.to.plot, 
                   reduction = "full.umap.sketch", ncol = 4))
  dev.off()
  
  # Spatial feature plots
  pdf("output/plots/top_marker_spatial.pdf", height = 20, width = 16)
  for (marker in markers.to.plot[1:8]) {  # Limit for space
    p <- SpatialFeaturePlot(merged.object, features = marker,
                           pt.size.factor = 2.5) +
      theme(legend.position = "right")
    print(p)
  }
  dev.off()
}
```

## 7. Spatial Analysis with BANKSY

### 7.1 BANKSY Setup and Analysis

```{r banksy_setup}
# Load BANKSY libraries
suppressPackageStartupMessages({
  library(Banksy)
  library(SeuratWrappers)
})

# Prepare data for BANKSY
merged.object <- NormalizeData(merged.object, assay = assay.name)
merged.object <- FindVariableFeatures(merged.object, nfeatures = 5000)

# Get consensus variable features across samples if multiple samples
if (length(unique(merged.object$orig.ident)) > 1) {
  # This would need custom implementation based on your specific needs
  # For now, use standard variable features
  consensus.var.feats <- VariableFeatures(merged.object)
} else {
  consensus.var.feats <- VariableFeatures(merged.object)
}

VariableFeatures(merged.object) <- consensus.var.feats
```

### 7.2 Prepare Spatial Coordinates

```{r banksy_coords}
# Extract spatial coordinates
locs.df.list <- lapply(unique(merged.object$orig.ident), function(sample_name) {
  # Get the image name for this sample
  available_images <- names(merged.object@images)
  image_name <- available_images[grepl(sample_name, available_images)][1]
  
  if (!is.na(image_name)) {
    loc.df <- GetTissueCoordinates(merged.object, image = image_name)
    return(loc.df)
  } else {
    # Fallback: use first available image
    loc.df <- GetTissueCoordinates(merged.object, image = available_images[1])
    return(loc.df)
  }
})

names(locs.df.list) <- unique(merged.object$orig.ident)
locs.df <- bind_rows(locs.df.list, .id = "orig.ident")
colnames(locs.df) <- c("orig.ident", "dimx", "dimy", "cell")

# Add coordinates to metadata
merged.object <- AddMetaData(merged.object, metadata = locs.df)
```

### 7.3 Run BANKSY Analysis

```{r banksy_analysis}
# Run BANKSY for different lambda values
for (lambda.val in banksy.lambda) {
  for (k.val in banksy.k) {
    cat("Running BANKSY with lambda =", lambda.val, "and k =", k.val, "\n")
    
    assay.banksy.name <- paste0("BANKSY_l", lambda.val, "_k", k.val)
    
    try({
      merged.object <- RunBanksy(merged.object,
        lambda = lambda.val,
        assay_name = assay.banksy.name,
        group = "orig.ident",
        verbose = FALSE,
        dimx = "dimx",
        dimy = "dimy",
        split.scale = TRUE,
        assay = assay.name,
        slot = "data",
        features = "variable",
        k_geom = k.val
      )
      
      # Run clustering on BANKSY features
      DefaultAssay(merged.object) <- assay.banksy.name
      merged.object <- RunPCA(merged.object, assay = assay.banksy.name, 
                             reduction.name = paste0("pca.banksy_l", lambda.val, "_k", k.val),
                             features = rownames(merged.object), npcs = 30)
      merged.object <- FindNeighbors(merged.object, 
                                   reduction = paste0("pca.banksy_l", lambda.val, "_k", k.val),
                                   dims = 1:30)
      merged.object <- FindClusters(merged.object, 
                                  cluster.name = paste0("banksy_l", lambda.val, "_k", k.val, "_res.", cluster.resolutions),
                                  resolution = cluster.resolutions)
      
      cat("BANKSY analysis completed for lambda =", lambda.val, "k =", k.val, "\n")
    })
  }
}

gc()  # Garbage collection
```

### 7.4 BANKSY Visualization

```{r banksy_viz, fig.width=20, fig.height=16}
# Visualize BANKSY results
pdf("output/plots/banksy_clustering_results.pdf", height = 16, width = 24)

for (lambda.val in banksy.lambda) {
  for (k.val in banksy.k) {
    for (res in cluster.resolutions[1:2]) {  # Limit to first two resolutions
      cluster.col <- paste0("banksy_l", lambda.val, "_k", k.val, "_res.", res)
      
      if (cluster.col %in% colnames(merged.object@meta.data)) {
        p <- SpatialDimPlot(merged.object, group.by = cluster.col,
                           label = TRUE, repel = TRUE, label.size = 3,
                           pt.size.factor = 2.5) +
          ggtitle(paste("BANKSY λ =", lambda.val, "k =", k.val, "res =", res)) +
          theme(legend.position = "none")
        print(p)
      }
    }
  }
}
dev.off()
```

## 8. Final Analysis and Export

### 8.1 Cluster Composition Analysis

```{r composition_analysis, fig.width=12, fig.height=8}
# Analyze cluster composition across samples
optimal.cluster.col <- paste0("sketch_snn_res.", optimal.res)

composition.df <- merged.object@meta.data %>%
  select(orig.ident, all_of(optimal.cluster.col)) %>%
  group_by(orig.ident, !!sym(optimal.cluster.col)) %>%
  summarise(counts = n(), .groups = "drop") %>%
  group_by(orig.ident) %>%
  mutate(percentage = counts / sum(counts) * 100)

# Composition bar plot
p1 <- ggplot(composition.df, aes(x = orig.ident, y = percentage, 
                                fill = !!sym(optimal.cluster.col))) +
  geom_bar(stat = "identity", position = "stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Cluster Composition Across Samples",
       x = "Sample", y = "Percentage", fill = "Cluster")

print(p1)

# Save composition data
write.csv(composition.df, "output/tables/cluster_composition.csv", row.names = FALSE)
```

### 8.2 Save Final Results

```{r save_results}
# Save final object
saveRDS(merged.object, file = "output/RDS-files/final_analysis_object.RDS")

# Create analysis summary
analysis.summary <- list(
  n_samples = length(unique(merged.object$orig.ident)),
  n_spots_total = ncol(merged.object),
  n_features = nrow(merged.object),
  resolutions_tested = cluster.resolutions,
  optimal_resolution = optimal.res,
  qc_thresholds = list(
    min_counts = min.counts,
    max_mt_percent = max.mt.percent
  ),
  banksy_parameters = list(
    lambda = banksy.lambda,
    k = banksy.k
  )
)

# Save summary as JSON
jsonlite::write_json(analysis.summary, "output/analysis_summary.json", 
                    pretty = TRUE, auto_unbox = TRUE)

cat("Analysis completed successfully!\n")
cat("Results saved in the 'output' directory\n")
```

## 9. Session Information

```{r session_info}
sessionInfo()
```

---

## Analysis Notes and Next Steps

### Key Outputs:
1. **RDS files**: Saved Seurat objects for future analysis
2. **Plots**: Comprehensive visualization of QC, clustering, and spatial patterns
3. **Tables**: Marker genes and cluster compositions
4. **Summary**: JSON file with analysis parameters and results

### Recommended Next Steps:
1. **Biological Interpretation**: Annotate clusters based on marker genes
2. **Functional Analysis**: Perform pathway enrichment on cluster markers
3. **Spatial Analysis**: Investigate spatial relationships between clusters
4. **Comparative Analysis**: Compare patterns across conditions/timepoints
5. **Integration**: Integrate with other omics data if available

### Customization Options:
- Adjust QC thresholds based on your dataset characteristics
- Modify clustering resolutions for optimal granularity
- Add custom gene sets for pathway analysis
- Include additional spatial analysis methods
- Implement batch correction if needed

This workflow provides a comprehensive foundation for Visium HD analysis that can be adapted to various experimental designs and research questions.